Gelation as condensation frustrated by hydrodynamics and mechanical tension

### 2016-11-25 gel_cond/trunk/manuscript4.tex
two columns, figures in the text

\documentclass[preprint,footinbib,amsmath,amssymb,superscriptaddress]{revtex4}
\usepackage{graphicx,amsmath}
\usepackage{color}
\begin{document}
\draft

@@

@@

%\subsection*{Observation of the initial stage}

@@

%\subsection*{Role of hydrodynamics}

@@

Consequently, long-lived elongated clusters have a higher probability to meet via either rotational or translational diffusion than compact spherical clusters. In Supplementary Figure~\ref{fig:effectiveVF} we show that indeed at all colloidal volume fraction percolation takes place when the effective volume fraction $\phi_\text{eff} = \sum\frac{4\pi}{3}R_g^3$ of the clusters reaches random close packing. Hydrodynamics explains why in a rather dilute regime we can observe immediate formation of elongated clusters and then their slow, hydrodynamically-assisted aggregation into a percolated structure.
We stress that this is a direct consequence of large-size disparity, which leads to the physical situation where discrete solid objects are floating in a continuum liquid.

%\subsection*{Network ageing}
In the above, we show that with a help of hydrodynamic interactions, particles first form
a transient gel even at very low volume fractions, and thus only later they increase the number of
nearest neighbours to minimize the energy of the structure. 
Thus, hydrodynamic interactions lead to the formation of gels that are very far from equilibrium and under a strong thermodynamic driving force 
towards more stable compact structures. 
Due to the connectivity of the network, the resulting transition from open to more compact networks can take place only through
the breaking of the bonds that have accumulated more stress. The process was called stress-driven ageing~\cite{Tanaka2007}, 
which is characteristic of viscoelastic phase separation \cite{tanaka2000viscoelastic}. 
As we saw in Fig.~\ref{fig:phasediag}c, the coarsening in these samples is indeed much slower than what is expected from ordinary spinodal decomposition dynamics \cite{onuki2002phase}, 
suggestive of mechanical frustration effects. The phase separation is arrested, but the system is still ageing slowly, a process sometimes leading to delayed sudden collapse of gels~\cite{poon1999delayed,Bartlett2012}, a catastrophic phenomenon from the industrial point of view. Previous numerical studies~\cite{Tanaka2007,colombo2014self,testard2014intermittent} have shown that the main coarsening mechanism is the rupture of strands in the network, a process that to our knowledge has not been followed experimentally at the particle level.
Below we provide the first experimental proof of this process.

@@

\clearpage
\begin{figure*}
	%\includegraphics[width=16cm]{figure1.pdf}
	\includegraphics{figs/cell_vs_cap2.pdf}
	\caption{\textbf{Reservoir cell.} \textbf{a,} Sketch of our experimental setup. The observation cell contains initially colloids, polymer and no salt. \textbf{b,} Confocal slice of a gel formed \textit{in situ} by our method ($\phi=25.5~\%$, $c_p=1.4$ g/$\ell$), 1 hour after gelation. \textbf{c,} Idem for a gel at the same state point formed \textit{ex situ} and immediately pumped into a capillary. \textbf{d,} Gelation observed by our method. Experimental coordinates are reconstructed and coloured by the number of particles in clusters in a typical sample close to the cluster-gel line ($\phi=7.5~\%$, $c_p=1$ g/$\ell$). Origin of time is the last frame before melting of Wigner crystal.
	%Phase diagram obtained in reservoir cell. Spinodal line is obtained from free volume theory. State points analysed in the text are circled. The state point of \textbf{b-c} is highlighted by a square.
	}
	\label{fig:cell_vs_cap}
\end{figure*}

\clearpage 
\begin{figure}
%\includegraphics[width=10cm]{figure2.pdf}
\includegraphics{figs/phasediag.pdf}
\caption{\textbf{Different regimes of gelation.} \textbf{a,} Phase diagram. Experimental points are categorised based on the final state obtained in reservoir cell. The spinodal line is obtained from free volume theory. State points analysed in the text are circled. The state point of Fig.~\ref{fig:cell_vs_cap}b-c is highlighted by a square. \textbf{b,} Comparison of system evolution in terms of largest cluster extent and of mean coordination number. By increasing density: $\phi=4.2,\,8,\,16,\,27~\%$, $c_p=1,\,1.5,\,1.2,\, 1$ g/$\ell$ for 
%\textcolor{red!40!yellow}
{$\blacklozenge$}, 
%\textcolor{red!80!yellow}
{$\circ$}, 
%\textcolor{red!80!black}
{\tiny$\blacksquare$} and 
%\textcolor{red!40!black}
{$\bullet$} respectively. \textbf{c,} Growth of the characteristic wave number for the same samples. 
The line is a scaling law for the intermediate coarsening regime for network-forming phase separation.}
\label{fig:phasediag}
\end{figure}

\clearpage
\begin{figure}
	%\includegraphics[width=10cm]{figure3.pdf}
	\includegraphics{figs/hydro.pdf}
	\caption{\textbf{Hydrodynamics} \textbf{a,} Probability of staying elongated for a triplet in a non-percolating sample ($\phi=4~\%$, $c_p=1$ g/$\ell$). The continuous line is the best exponential fit of characteristic time $27\tau_B$. Inset: probability distribution of radii of gyration of triplets. \textbf{b,} Evolution of the aspect ratios of clusters of 4 particles and more in the same sample (dashed lines) and in a percolating sample ($\phi=8~\%$, $c_p=1.5$ g/$\ell$, continuous lines). \textbf{c,} Bond angle distribution relative to existing bonds (gray), to a future bond (red) or to a future bond involving an isolated particle (blue) obtained in the percolating sample. Future bonds are shifted to smaller angles, whereas gas adsorption takes place from larger angles. Insets sketch both cases, with present bonds drawn thick and future bonds drawn dotted.}
	\label{fig:hydro}
\end{figure}

\clearpage
\begin{figure*}
%\includegraphics[width=16cm]{figure4.pdf}
\includegraphics{figs/breaking.pdf}
\caption{\textbf{Mechanical tension drives coarsening} 
\textbf{a,} Reconstruction from experimental coordinates ($\phi=29~\%$, $c_p=0.7$ g/$\ell$) of coarsening process geometry (top) and topology (bottom). Particles are drawn to scale and coloured by $q_2$ from blue (low) to red (high). The meshed surface is a Gaussian coarse-graining of the network pattern. The red line indicates the shortest on-graph path between the two particles of interest. 
\textbf{b,} Number of bond breaking function of the future on-graph distance $\Lambda$ (after percolation) for the three percolating state points of Fig. 2a (same colours and markers).
%three state points of increasing density: $\phi=8,\,16,\,27~\%$, $c_p=1.5,\,1.2,\,1$g/$\ell$ for \textcolor{red!80!yellow}{$\circ$}, \textcolor{red!80!black}{\tiny$\blacksquare$} and \textcolor{red!40!black}{$\bullet$} respectively.
\textbf{c,} Probability for a bond to break during $3\tau_B$ function of bond-centred $q_2$ for all events (blue) and for events with $\Lambda>1$ (thick red) in the same sample as \textbf{a}.
\textbf{d,} Evolution of the topologically relevant bond breaking probability ($\Lambda>3$) for the same state points as in b. Vertical lines indicate the respective percolation times.
}
\label{fig:breaking}
\end{figure*}

>>>>

\documentclass[footinbib,amsmath,amssymb,superscriptaddress,twocolumn]{revtex4}
\usepackage{graphicx,amsmath}
\usepackage{color}
\begin{document}

@@

\begin{figure*}
	%\includegraphics[width=16cm]{figure1.pdf}
	\includegraphics{figs/cell_vs_cap2.pdf}
	\caption{\textbf{Reservoir cell.} \textbf{a,} Sketch of our experimental setup. The observation cell contains initially colloids, polymer and no salt. \textbf{b,} Confocal slice of a gel formed \textit{in situ} by our method ($\phi=25.5~\%$, $c_p=1.4$ g/$\ell$), 1 hour after gelation. \textbf{c,} Idem for a gel at the same state point formed \textit{ex situ} and immediately pumped into a capillary. \textbf{d,} Gelation observed by our method. Experimental coordinates are reconstructed and coloured by the number of particles in clusters in a typical sample close to the cluster-gel line ($\phi=7.5~\%$, $c_p=1$ g/$\ell$). Origin of time is the last frame before melting of Wigner crystal.
	%Phase diagram obtained in reservoir cell. Spinodal line is obtained from free volume theory. State points analysed in the text are circled. The state point of \textbf{b-c} is highlighted by a square.
	}
	\label{fig:cell_vs_cap}
\end{figure*}

@@

%\subsection*{Observation of the initial stage}
\begin{figure}
%\includegraphics[width=10cm]{figure2.pdf}
\includegraphics{figs/phasediag.pdf}
\caption{\textbf{Different regimes of gelation.} \textbf{a,} Phase diagram. Experimental points are categorised based on the final state obtained in reservoir cell. The spinodal line is obtained from free volume theory. State points analysed in the text are circled. The state point of Fig.~\ref{fig:cell_vs_cap}b-c is highlighted by a square. \textbf{b,} Comparison of system evolution in terms of largest cluster extent and of mean coordination number. By increasing density: $\phi=4.2,\,8,\,16,\,27~\%$, $c_p=1,\,1.5,\,1.2,\, 1$ g/$\ell$ for 
%\textcolor{red!40!yellow}
{$\blacklozenge$}, 
%\textcolor{red!80!yellow}
{$\circ$}, 
%\textcolor{red!80!black}
{\tiny$\blacksquare$} and 
%\textcolor{red!40!black}
{$\bullet$} respectively. \textbf{c,} Growth of the characteristic wave number for the same samples. 
The line is a scaling law for the intermediate coarsening regime for network-forming phase separation.}
\label{fig:phasediag}
\end{figure}

@@

%\subsection*{Role of hydrodynamics}
\begin{figure}
	%\includegraphics[width=10cm]{figure3.pdf}
	\includegraphics{figs/hydro.pdf}
	\caption{\textbf{Hydrodynamics} \textbf{a,} Probability of staying elongated for a triplet in a non-percolating sample ($\phi=4~\%$, $c_p=1$ g/$\ell$). The continuous line is the best exponential fit of characteristic time $27\tau_B$. Inset: probability distribution of radii of gyration of triplets. \textbf{b,} Evolution of the aspect ratios of clusters of 4 particles and more in the same sample (dashed lines) and in a percolating sample ($\phi=8~\%$, $c_p=1.5$ g/$\ell$, continuous lines). \textbf{c,} Bond angle distribution relative to existing bonds (gray), to a future bond (red) or to a future bond involving an isolated particle (blue) obtained in the percolating sample. Future bonds are shifted to smaller angles, whereas gas adsorption takes place from larger angles. Insets sketch both cases, with present bonds drawn thick and future bonds drawn dotted.}
	\label{fig:hydro}
\end{figure}

@@

Consequently, long-lived elongated clusters have a higher probability to meet via either rotational or translational diffusion than compact spherical clusters. 
%In Supplementary Figure~\ref{fig:effectiveVF} we show that indeed at all colloidal volume fraction percolation takes place when the effective volume fraction $\phi_\text{eff} = \sum\frac{4\pi}{3}R_g^3$ of the clusters reaches random close packing. 
Hydrodynamics explains why in a rather dilute regime we can observe immediate formation of elongated clusters and then their slow, hydrodynamically-assisted aggregation into a percolated structure.
We stress that this is a direct consequence of large-size disparity, which leads to the physical situation where discrete solid objects are floating in a continuum liquid.

\begin{figure*}
%\includegraphics[width=16cm]{figure4.pdf}
\includegraphics{figs/breaking.pdf}
\caption{\textbf{Mechanical tension drives coarsening} 
\textbf{a,} Reconstruction from experimental coordinates ($\phi=29~\%$, $c_p=0.7$ g/$\ell$) of coarsening process geometry (top) and topology (bottom). Particles are drawn to scale and coloured by $q_2$ from blue (low) to red (high). The meshed surface is a Gaussian coarse-graining of the network pattern. The red line indicates the shortest on-graph path between the two particles of interest. 
\textbf{b,} Number of bond breaking function of the future on-graph distance $\Lambda$ (after percolation) for the three percolating state points of Fig. 2a (same colours and markers).
%three state points of increasing density: $\phi=8,\,16,\,27~\%$, $c_p=1.5,\,1.2,\,1$g/$\ell$ for \textcolor{red!80!yellow}{$\circ$}, \textcolor{red!80!black}{\tiny$\blacksquare$} and \textcolor{red!40!black}{$\bullet$} respectively.
\textbf{c,} Probability for a bond to break during $3\tau_B$ function of bond-centred $q_2$ for all events (blue) and for events with $\Lambda>1$ (thick red) in the same sample as \textbf{a}.
\textbf{d,} Evolution of the topologically relevant bond breaking probability ($\Lambda>3$) for the same state points as in b. Vertical lines indicate the respective percolation times.
}
\label{fig:breaking}
\end{figure*}

%\subsection*{Network ageing}
In denser samples, hydrodynamics also helps create elongated structures, but in that case these structures immediately connect and percolate. The increase of the number of
nearest neighbours to minimize the energy of the structure comes only later.
%In the above, we show that with a help of hydrodynamic interactions, particles first form
%a transient gel even at very low volume fractions, and thus only later they increase the number of
%nearest neighbours to minimize the energy of the structure. 
Thus, hydrodynamic interactions lead to the formation of gels that are very far from equilibrium and under a strong thermodynamic driving force 
towards more stable compact structures. 
Due to the connectivity of the network, the resulting transition from open to more compact networks can take place only through
the breaking of the bonds that have accumulated more stress. The process was called stress-driven ageing~\cite{Tanaka2007}, 
which is characteristic of viscoelastic phase separation \cite{tanaka2000viscoelastic}. 
As we saw in Fig.~\ref{fig:phasediag}c, the coarsening in these samples is indeed much slower than what is expected from ordinary spinodal decomposition dynamics \cite{onuki2002phase}, 
suggestive of mechanical frustration effects. The phase separation is arrested, but the system is still ageing slowly, a process sometimes leading to delayed sudden collapse of gels~\cite{poon1999delayed,Bartlett2012}, a catastrophic phenomenon from the industrial point of view. Previous numerical studies~\cite{Tanaka2007,colombo2014self,testard2014intermittent} have shown that the main coarsening mechanism is the rupture of strands in the network, a process that to our knowledge has not been followed experimentally at the particle level.
Below we provide the first experimental proof of this process.

@@

\clearpage

\clearpage
